\problem{10}
The isentropic compressibility is defined by $\kappa_S \equiv -\frac{1}{V}\pdc{V}{p}{S}$.
Show that $\kappa_S$ relates to the isobaric and isochoric heat capacities as follows:
\begin{align*}
C_V &= \frac{TV\alpha^2 \kappa_S}{(\kappa_T - \kappa_S)\kappa_T}  \\
C_p &= \frac{TV\alpha^2}{\kappa_T - \kappa_S} .
\end{align*}
For ideal gas, find an expression that relates $\kappa_S$ to $C_p$, $p$ and $V$.

\bigskip
\problem{6}
The equation of state for non-ideal fluids may be analyzed by studying
the compressibility factor $Z$ defined by $Z\equiv pv/(RT)$,
where $v$ is the molecular molar volume $V/N$.
The virial expansion for $Z$ is defined as
$$Z = 1 + \frac{B_2}{v} + \frac{B_3}{v^2} + \cdots$$
where $B_2$, $B_3$, $\cdots$ are called the second, the third,
$\ldots$ virial coefficients, which are determined by the inter-molecular interactions.

\smallskip
\subp
Find the second and the third virial coefficents for a van der Waals fluid with an equation of state 
$p = RT/(v - b) - a / v^2$.

\smallskip
\subp
The second virial coefficient relates to the pairwise inter-molecular potential $u(r)$ by
$$ B_2 = 2\pi \NA \int_0^\infty\left[1 - \exp\left(-\frac{u(r)}{\kT}\right)\right] r^2 dr. $$
%In the ``hard-sphere'' model, the interaction potential between two molecules is given by
%\begin{equation*}
%\Gamma(r) = \left\{\begin{aligned}
%     0,\quad& r\ge d, \\
%\infty,\quad& 0 < r < d.
%\end{aligned}\right.
%\end{equation*}
At short distance, the inter-molecular potential is dominated by repulsion.
Assume $u(r)$ is given by $u(r) = c\, r^{-n} \, (n>3)$, in which $c$ is a positive constant.
Show that $B_2$ is given by
\begin{equation*}
B_2 = \frac{2\pi \NA}{3} \left(\frac{c}{\kT}\right)^{3/n} \Gamma(1-3/n),
\end{equation*}
where the $\Gamma$ function is defined by
$\Gamma(z) \equiv \int_{0}^\infty\!dx\, \ee^{-x} x^{z-1} $.
{\bf Hint}: {\sl try integration by part}.

\bigskip
\problem{15}
Reduce the following five derivatives in terms of standard quantities,
$C_p$, $\alpha$ and $\kappa_T$.

\begin{equation*}
\pdc{S}{V}{p}, \quad
\pdc{V}{T}{S}, \quad
\pdc{H}{V}{T}, \quad
\pdc{V}{p}{H}, \quad
\pdc{V}{S}{H}
\end{equation*}

\bigskip
\problem{20}
Consider a gas that is at a high pressure $p_1$ at temperature $T_1$
that is throttled through a valve to a lower pressure $p_2$.
The throttle valve is insulated, so no heat is passed to the gas as it goes through the valve.
No work is done in the process.
Describe the valve as an open system with inlet at pressure $p_1$ and outlet at pressure $p_2$.

\rightline{See next page for questions.}
\vfil\eject

\ \medskip
\subp
Given that the valve is an open system,
what is the relationship between the molar enthalpy of the inlet stream
$H_{1}$ and outlet stream $H_{2}$?

\smallskip
\subp
To calculate the change in temperature for this process, define the Joule-Thomson coefficient
$$\mu = \left( \frac{\partial T}{\partial p} \right)_{H} $$
Explain in a couple of sentences how $\mu$ can be use to find the outlet temperature $T_{2}$ from $p_{1}$, $T_{1}$, $p_{2}$.

\smallskip
\subp
Derive an expression for $\mu$ is terms of standard thermodynamic variables
and measurables $C_{p}$, $\alpha$, and $\kappa_T$.

\smallskip
\subp
Show that the Joule-Thomson coefficient is zero for an ideal gas.

\smallskip
\subp
Consider the virial equation of state
$$ \frac{pv}{RT} = 1 + \frac{B(T)}{v} $$
Derive an expression for $\mu$ in terms of $T$, $B$, and $C_{p}$.

\smallskip
\subp
Explain in a couple of sentences the conditions where $\mu$ is positive, negative, and zero.
In your explanation, clearly state the thermodynamic properties that
are constrained to be positive due to thermodynamic stability.

